Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-16 paper.qmd
Replaced 4-D with notation

```{r}
#| label: fig-two-curvy-true-proj
#| fig-cap: "Two \\gD{} projections from $4\\text{-}D$, for the two nonlinear clusters data (blue points) and true model (black lines) are shown. The top cluster in (a) shows uniformly distributed data along a C-shape, while in the bottom cluster, the data is more concentrated in one corner of the curvy shape. Video of the langevitour animations are available at <>." 
#| fig-pos: H
#| fig-width: 10
#| fig-height: 5
#| eval: false

@@

```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the fitted model into \\pD{}. Starting from a default tSNE layout (a) of nonlinear clusters data ($n =  2000$ and $p = 4$), from which the \\gD{} wireframe is constructed (a) and then lifted to $4\\text{-}D$ (b, c). The fit is reasonbly tight with the data, although it does not fully spread the full width. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-)~parameter choice. Video of the langevitour animation is available at <>."
#| fig-width: 8
#| fig-height: 2

@@

<!-- Sparse end with dense corner to me is very hard to understand - sparse and dense are pretty much antonyms -->
```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "The $4\\text{-}D$ model error in \\gD{} layout. Color indicates error ($e_{hj}$), dark color indicating high error and light indicates low error. Most large errors are distributed near the sparse end of the nonlinear cluster and most small errors are distributed near the dense corners."
#| out-width: 80%
#| fig-pos: H

@@

```{r}
#| label: fig-five-gau-projs
#| fig-height: 10
#| fig-width: 15
#| fig-pos: H
#| fig-cap: "NLDR's organise points in the \\gD{} layout in different ways, possibly misleadingly, illustrated using three layouts: (a) tSNE, (b) UMAP, (c) PaCMAP. The data has five Gaussian cluster in $4\\text{-}D$. The bottom row of plots shows a \\gD{} projection from a tour on $4\\text{-}D$ revealing the differences generated by the layouts on the model fits.  We would expect the model fit to be like that in (a2) where it is distinctly is separate for each cluster but like a hairball in each. This would indicate the distinct clusters, each being fully $4\\text{-}D$. With (c2), the curiousity is that the model is a \\gD{} pancake shape in $4\\text{-}D$, indicating that there is some ordering of points done by PaCMAP, posisbly along some principal component axes.  Videos of the langevitour animations are available at <https://youtu.be/I-kxCwVfqiQ>, <https://youtu.be/gD1P01FUPyU>, and <https://youtu.be/MxJ_srOFQNk> respectively."

@@

```{r}
#| label: fig-one-dens_clust-error
#| fig-width: 15
#| fig-height: 10
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "The sparse end of the structure shows a high model error, while the dense end shows a low model error. The tSNE layout of the C-shaped structure, with dense points (a1-a3) are colored according to their $4\\text{-}D$ model error. Darker colors represent a high error, while lighter colors indicate a low error. a2, a3 and b2, b3 present the results from linked plots that brushed the high $4\\text{-}D$ error points in \\gD{} and $4\\text{-}D$. It helps in identifying the high $4\\text{-}D$ model errors that occur due to the sparse end. Videos of the linked plots are available at <>."

>>>>

```{r}
#| label: fig-two-curvy-true-proj
#| fig-cap: "Two \\gD{} projections from \\fD{}, for the two nonlinear clusters data (blue points) and true model (black lines) are shown. The top cluster in (a) shows uniformly distributed data along a C-shape, while in the bottom cluster, the data is more concentrated in one corner of the curvy shape. Video of the langevitour animations are available at <>." 
#| fig-pos: H
#| fig-width: 10
#| fig-height: 5
#| eval: false

@@

```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the fitted model into \\pD{}. Starting from a default tSNE layout (a) of nonlinear clusters data ($n =  2000$ and $p = 4$), from which the \\gD{} wireframe is constructed (a) and then lifted to \\fD{} (b, c). The fit is reasonbly tight with the data, although it does not fully spread the full width. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-)~parameter choice. Video of the langevitour animation is available at <>."
#| fig-width: 8
#| fig-height: 2

@@

<!-- Sparse end with dense corner to me is very hard to understand - sparse and dense are pretty much antonyms -->
```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "The \\fD{} model error in \\gD{} layout. Color indicates error ($e_{hj}$), dark color indicating high error and light indicates low error. Most large errors are distributed near the sparse end of the nonlinear cluster and most small errors are distributed near the dense corners."
#| out-width: 80%
#| fig-pos: H

@@

```{r}
#| label: fig-five-gau-projs
#| fig-height: 10
#| fig-width: 15
#| fig-pos: H
#| fig-cap: "NLDR's organise points in the \\gD{} layout in different ways, possibly misleadingly, illustrated using three layouts: (a) tSNE, (b) UMAP, (c) PaCMAP. The data has five Gaussian cluster in \\fD{}. The bottom row of plots shows a \\gD{} projection from a tour on \\fD{} revealing the differences generated by the layouts on the model fits.  We would expect the model fit to be like that in (a2) where it is distinctly is separate for each cluster but like a hairball in each. This would indicate the distinct clusters, each being fully \\fD{}. With (c2), the curiousity is that the model is a \\gD{} pancake shape in \\fD{}, indicating that there is some ordering of points done by PaCMAP, posisbly along some principal component axes.  Videos of the langevitour animations are available at <https://youtu.be/I-kxCwVfqiQ>, <https://youtu.be/gD1P01FUPyU>, and <https://youtu.be/MxJ_srOFQNk> respectively."

@@

```{r}
#| label: fig-one-dens_clust-error
#| fig-width: 15
#| fig-height: 10
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "The sparse end of the structure shows a high model error, while the dense end shows a low model error. The tSNE layout of the C-shaped structure, with dense points (a1-a3) are colored according to their \\fD{} model error. Darker colors represent a high error, while lighter colors indicate a low error. a2, a3 and b2, b3 present the results from linked plots that brushed the high \\fD{} error points in \\gD{} and \\fD{}. It helps in identifying the high \\fD{} model errors that occur due to the sparse end. The link to the video of the linked plots is available in @tbl-links."
